Extended Data

### 2021-04-29 doc/manuscript/Dow_manuscript_Nature.Rmd
writing + clean-up

**Authors (not yet complete, just alphabetical for now--final order TBD):**

@@

Grant Harley

Justin Maxwell

Ian McGregor^1,6^

@@

Sean McMahon^7,9^

@@

Alan Tepley^1,8^

Kristina J. Anderson-Teixeira^1,9^*

@@

6. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA

7. Smithsonian Environmental Research Center, Edgewater, MD, USA

8. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

9. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

@@

(*word limit 200. italic=tentative*)

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[*@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016*; @zhou_explaining_2016;  @fu_maximum_2019].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig. `r tree_rings`; `r EDF_map`).
In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only #% of records: # of # ring-porous and # or # diffuse-porous species-site combinations.
In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.

@@

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress driven by increased vapor pressure deficit[**REFS**].
It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, effectively reducing drought sensitivity, but this hypothesis remains to be evaluated.

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that the extra C does not go into woody growth, raising the question of where this C goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013]. 
Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, and that this dynamic is likely complex[@helcoski_growing_2019].
Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
A study examining long-term trends in growing season length and NEE[@keenan_net_2014] -- as opposed to their interannual variation -- showed increasing trends in both within the temperate deciduous biome, suggesting that the C that is not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
In addition to the existence of temporal lags between C uptake and release, which introduce a lot of uncertainty[@trumbore_quantifying_2002], trees are likely to  acclimate *to some extent* to warming temperatures[**REFS?**], and on even longer time scales, we are likely to see changes in forest composition and adaptation.
If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
*Future research will be important to sortitng this out.*

*(KAT edits end here)*

Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
Warm spring temperatures result in earlier leaf-out (Fig. `r DOY_timing` , REFS), an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs. `r schematic_summary`- `r pheno_Tsensitivity`).
Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs. `r schematic_summary`- `r tree_ring`), presumably because stem diameter increases are curtailed by other factors such as...., supporting the concept of *C sink limitation* [@zani_increased_2020].

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models (Figs. `r schematic_summary`, `r pheno_Tsensitivity`, `r tree_rings`)[@zani_increased_2020].
Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.

@@

From Neil: @zhang_drought_2021, @zhang_extended_2020

@@

The timing of tree growth has profound impacts on ecosystem interactions, as well as dictates periods of highest drought sensitivity in temperate forests (Foster, 2014; D’Orangeville et al. in review).

@@

### Leaf phenology
description here: https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description

MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019]

@@

The tree-ring records from our focal sites were complemented with a much larger collection spanning **#** deciduous and mixed forest sites in Eastern North America.
Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).

@@

## Determining critical temperature windows {-#Determining}

The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window. 
Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r  EDT_mean_growth_parameters`).
*The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic_summary`), because we were interested in the influence of temperature on the initiation of woody growth. 
We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019]. Determining the climate variables that most strongly control all parameters is beyond the scope of this analysis.

## Hypothesis testing {-#Hypothesis}

(*This section needs more detail.*)

>>>>

**Authors:**

@@

Grant Harley^5^

Justin Maxwell^6^

Ian McGregor^1,7^

@@

Sean McMahon^8,10^

@@

Alan Tepley^1,9^

Kristina J. Anderson-Teixeira^1,10^*

@@

6. Department of Geography, Indiana University, Bloomington, Indiana, USA

7. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA

8. Smithsonian Environmental Research Center, Edgewater, MD, USA

9. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

10. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

@@

(*word limit 200 - this is currently way too long.*)

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

The patterns observed at our focal sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`)[@dorangeville_drought_2018].
In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only #% of records: # of # ring-porous and # or # diffuse-porous species-site combinations.
In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.

@@

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season, potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019], but this hypothesis remains to be evaluated.

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that the extra C does not go into woody growth, raising the question of where this C goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013]. 
Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome  showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
In addition to the existence of temporal lags between C uptake and release, which introduce a lot of uncertainty[@trumbore_quantifying_2002], trees are likely to  acclimate *to some extent* to warming temperatures[**REFS?**], and on even longer time scales, we are likely to see changes in forest composition and adaptation.
If we look across spatial gradients where the latter have had time to play out, we see that longer growing seasons are correlated with greater tree growth[**REF?**], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
*Future research will be important to sorting this out.*

Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
Warm spring temperatures result in earlier leaf-out (Fig. `r DOY_timing`)[@jeong_phenology_2011], an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs. `r schematic_summary`- `r pheno_Tsensitivity`).
Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs. `r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@dorangeville_drought_2018;@helcoski_growing_2019].

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the sustained increase in carbon sequestration that is anticipated in most models (Fig. `r schematic_summary`)[@zani_increased_2020].
Because of this, temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that these models are underestimating future overall atmospheric CO~2~.
(*see [GitHub issue 56 on representation in models](https://github.com/EcoClimLab/growth_phenology/issues/56)*)

@@

From Neil: @zhang_drought_2021; @zhang_extended_2020

@@

@@

### Leaf phenology

Leaf phenology data for the years ### - ### were extracted for SCBI and Harvard forest from the MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019].

description here: https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description

@@

The tree-ring records from our focal sites were complemented with a much larger collection spanning `r n_cores_sites-2` deciduous and mixed forest sites in Eastern North America[@dorangeville_drought_2018] (**OTHER REFS??**).
Again, species were limited to broadleaf deciduous species with clearly defined xylem porosity (i.e., excluding semi-ring porous).

@@

## Analyses
### Determining critical temperature windows {-#Determining}

The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and $\Delta AIC$ of each window. 
Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r  EDT_mean_growth_parameters`).
*The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic_summary`), because we were interested in the influence of temperature on the initiation of woody growth. 
We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019]. Determining the climate variables that most strongly control all parameters is beyond the scope of this analysis.

### Hypothesis testing {-#Hypothesis}

(*This section needs work/ more detail.*)
